---
title: FAB p-values for small areas
author: Peter Hoff 
date: 2019-07-23
---

----

## Summary 
Adaptive FAB $p$-values are constructed  for evaluating the mean test scores of 10th graders in a population of high schools. Information is shared across schools using a Fay-Herriot random effects model. This document serves as the replication code for the example in Section 3.2 of the article
["Smaller $p$-values via indirect information"](https://arxiv.org/abs/1907.12589) (Hoff 2019).




## Data
The data for this example are from the 2002 Educational Longitudinal Survey, data from which can 
be obained at https://nces.ed.gov/surveys/els2002/. A subset of the data needed to 
replicate the example is available on my website and can be loaded using the following command:
```{r}
load(url("http://www2.stat.duke.edu/~pdh10/Datasets/els.RData")) 
```

Now take a look at all the variables:
```{r}
els[1:10,]
```



Separate out student-level characteristics from school-level characteristics:
```{r} 
y<-els$rscore 
group<-els$school 
groups<-sort(unique(group))  

W<-as.matrix(els[,c("flp","urban","rural","private","catholic","rm","rs","rw","enrollment") ] )  
X<-apply(W,2,function(x){ tapply(x,group,"mean") } ) 
```

Here, `y` is the student-level reading score, and `group` 
is the school id for each student. That is, 
`group[i]` is the school of student with score `y[i]`. 
The matrix `X` is a matrix of school-level variables, and the 
matrix `W` is the same data except with the variable for each school replicated for each student in the school. 


## UMPU $p$-values 
Let $\theta_j$ be the "true" mean test score in school $j$. 
The standard two-sided $p$-value for testing 
$H_j:\theta_j=50$  is 
based on the $t$-statistic $t_j = \sqrt{n_j}( \bar y_j - 50)/\hat\sigma_j$. 
```{r}
theta0<-50
YBAR<-tapply(y,group,mean) - theta0
S2<-tapply(y,group,var)
N<-c(table(group))
T<-YBAR/sqrt(S2/N)
pU<-c(2*pt(-abs(T),N-1))
```



## FH linking model
To compute the FAB $p$-values, we need to first obtain 
indirect information about each $\theta_j$ using data from schools other than 
school $j$. 
We quantify this indirect information using the 
following linking model for school-specific means:
\[
 \theta_j =  \beta^\top x_j + \tau \gamma_j 
\]
where $\beta$ and $\tau$ are unknown, 
$x_j$ is a vector of observed characteristics of school $j$, 
and $\gamma_1,\ldots, \gamma_p \sim$ are i.i.d. standard normal 
random variables. For each school $j$, we get an estimate of these 
parameters using the data from schools other than $j$, along with 
an estimate of the within school sampling variance $\sigma^2$. This is 
done using the `lmer` command in the `lme4` package:
```{r,cache=TRUE}
library(lme4)
BETA<-TAU<-SIGMA<-NULL
for(i in 1:length(groups)){

  g<-groups[i]
  iy<-y[group!=g]
  iW<-W[group!=g,]
  igroup<-group[group!=g]

  ifit<-lmer(iy ~ iW + (1|igroup) )

  beta<-fixef(ifit)
  tau<-sqrt(unlist(VarCorr(ifit)))
  sigma<-attr(VarCorr(ifit), "sc")

  BETA<-rbind(BETA,beta)
  TAU<-c(TAU,tau)
  SIGMA<-c(SIGMA,sigma)

  #cat(g,"\n") 
}
```
Note that the $j$th elements of `BETA`, `TAU` and `SIGMA` are estimated 
WITHOUT data from school $j$, and so are independent of the data in school 
$j$. 


## FAB $p$-values 

Based on these linking model parameter estimates we compute an 
indirect/prior mean $\theta_j$: 
Given school-level covariates $x_j$, the fitted predicted value for 
$\theta_j$ based on the linking model is 
$\hat\beta^\top x_j$: 
```{r}
ETHETA<-apply( BETA*cbind(1,X) ,1,sum) - theta0
``` 
Here we have centered things around the hypothesized mean value of 
$\theta_0=50$. 

The FAB $p$-value for school $j$ is 
$1-|F( T_j + \tilde b_j ) - F(-T_j)|$, 
where $T_j$ is the $t$ statistic for school $j$, 
$b_j = 2 \tilde \mu (\tilde \sigma/\sqrt{n_j})/\tilde\tau^2$, 
and $F$ is the CDF of the appropriate $T$-distribution. 
```{r} 
pF<- c(1-abs( pt(T+2*ETHETA*(SIGMA/sqrt(N))/(TAU^2),N-1)  - pt(-T,N-1) ) )
```

Save some results:
```{r}
resultsFHmodel<-list(T=T,YBAR=YBAR,S2=S2,N=N,pU=pU,ETHETA=ETHETA,TAU=TAU,SIGMA=SIGMA,pF=pF)
save(resultsFHmodel,file="resultsFHmodel.RData")
```

## Summary of results  

```{r}
sum(pF<=.05)
sum(pU<=.05)
mean(pF<=pU)
```


```{r}
par(mfrow=c(1,3),mar=c(3,3,1,1),mgp=c(1.75,.75,0))

plot(resultsFHmodel$ETHETA,resultsFHmodel$YBAR,
     xlab="indirect estimate of school-level mean",
     ylab="direct estimate of school-level mean")
abline(0,1,col="gray")

pU<-resultsFHmodel$pU
pF<-resultsFHmodel$pF

plot(pU,pF,xlab="UMPU p-value",ylab="FAB p-value")
abline(0,1,col="gray")


plot(sort(pU),type="l",xlim=c(0,350),ylim=c(0,.06),lwd=2,col="pink",
          xlab="rank",ylab="p-value")
lines(sort(pF),col="lightblue",lwd=2)
abline(h=.05,col="gray")
legend(5,.045,lwd=c(2,2),col=c("pink","lightblue"),legend=c("UMPU","FAB"),
       bty="n")
```



