---
title: Asymptotically uniform FAB $p$-values 
author: Peter Hoff
date: 2019-07-23
---

----

## Summary 
This is a simulation study in which 
adaptive FAB $p$-values are constructed 
for logistic regression coefficients. 
This document serves as the replication code for the example in Section 4.3 of the article
["Smaller $p$-values via indirect information"](https://arxiv.org/abs/XXXX) (Hoff 2019).


---



## Linking model 

The linking model we consider here is that the logistic regression coefficients 
$\theta_1,\ldots, \theta_p$ are i.i.d. from a mixture of a point-mass distribution 
at zero and a normal distribution with some mean and variance. 

Here is a function that estimates the parameters in the linking model. 
More generally, it obtains MLEs of $\mu$ and $\tau^2$ 
under the model 
\[
\begin{align*} 
y & \sim N_p( \theta , \sigma^2 I) \\
\theta_1,\ldots, \theta_p &\sim i.i.d\ (1-\pi) \delta_0(\theta) + \text{dnorm}(\theta, \mu,\tau) 
\end{align*}
\]

```{r}
fitMM<-function(y,s2){

  obj<-function(gam){
    pi<-gam[1] ; mu<-gam[2] ; t2<-gam[3]
    -sum( log( pi*dnorm(y,mu,sqrt(t2+s2)) + (1-pi)*dnorm(y,0,sqrt(s2))) )
  }

  mut2<-optim( c(.9,mean(y),var(y)),obj,method="L-BFGS-B",
               lower=c(.001,-Inf,.001),upper=c(.999,Inf,Inf))$par[2:3]
  names(mut2)<-c("mu","t2")
  mut2
}
```

For this example, for each $\theta_j$ 
we will plug-in an estimate $\hat\theta_{-j}$ of 
the vector $\theta_{-j}$ using data that is independent (asymptotically)
of $\hat\theta_j$. 


Here is the simulation:
```{r,cache=TRUE}
set.seed(1)
p<-30
theta0<-rep(c(3,0),times=c(p/2,p/2))
nvals<-c(200,400,800,1600) 
nsim<-5000

PVALS<-array(dim=c(nsim,p,2,length(nvals))) 

par(mfrow=c(4,3),mar=c(3,3,1,1),mgp=c(1.75,.75,0)) 

for(k in 1:length(nvals)){

  n<-nvals[k] 

  PF<-PW<-NULL
  for(s in 1:nsim){ 

    X<-matrix(rnorm(n*p),n,p) 
    lp<-X%*%theta0/sqrt(n)
    y<-rbinom(n,1,1/(1+exp(-lp))) 
    fit<-glm(y~-1+X,family="binomial")

    theta<-fit$coef
    Sigma<-summary(fit)$cov.scaled
    s2<-diag(Sigma)  

    MUT2<-NULL
    for(j in 1:p){
      G<-MASS::Null(Sigma[,j])
      MUT2<-rbind(MUT2,fitMM(  (G%*%(t(G)%*%theta))[-j],s2[-j]) )
    }

    z<-theta/sqrt(s2)
    pW<-1-abs( pnorm(z)-pnorm(-z))
    pF<-1-abs( pnorm(z+2*sqrt(s2)*MUT2[,1]/MUT2[,2]) - pnorm(-z) ) 
    PF<-rbind(PF,pF) ; PW<-rbind(PW,pW)  
  } 

  PVALS[,,1,k]<-PW 
  PVALS[,,2,k]<-PF 


  hist(PF[,theta0!=0],col=rgb(.3,.3,.3,.5),prob=TRUE) 
  hist(PW[,theta0!=0],col=rgb(.8,.2,.5,.5),add=TRUE,prob=TRUE) 

  hist(PF[,theta0==0]) 
  hist(PW[,theta0==0]) 
 
}
```

Here we save the results:
```{r}
dimnames(PVALS)[[3]]<-c("pW","pF") 
dimnames(PVALS)[[4]]<-nvals  
resultsLogistic <-PVALS
save(resultsLogistic,file="resultsLogistic.RData") 
```

The results are in $30\times 5000 \times 2 \times 4$ array. 
The first dimension corresponds to the regression coefficients, 
the first 15 of which were $3/\sqrt{n}$ and the last 15 of which were zero. 
The second dimension corresponds to the simulation number (1 to 5000). 
The third dimension is the $p$-value type, and the fourth dimension 
is the sample size. 


Here are the plots from the article:
```{r}
PVALS<-resultsLogistic
p<-dim(PVALS)[2]
theta0<-rep(c(3,0),times=c(p/2,p/2))

par(mfrow=c(4,3),mar=c(2.75,3.2,0,1),mgp=c(1.75,.75,0))
for(k in 1:dim(PVALS)[4]){

  PW<-PVALS[,,1,k]
  PF<-PVALS[,,2,k]

  xlab<-1*(k==dim(PVALS)[4]) + 1

  hist(PF[,theta0!=0],col=rgb(.1,.4,.6,.5),prob=TRUE,main="",
       xlab=c("","non-null p-values")[xlab],ylab="",xaxt=c("n","s")[xlab] )
  hist(PW[,theta0!=0],col=rgb(.9,.3,.3,.5),prob=TRUE,add=TRUE)
  mtext(paste0("n=",dimnames(PVALS)[[4]][k]),2,line=2,cex=.85)

  if(k==1){
    legend(.5,7,legend=c("Wald","FAB"),pch=15,col=c(rgb(.9,.3,.3,.5),rgb(.1,.4,.6,.5)),bty="n")
  }

  hist(PW[,theta0==0],prob=TRUE,col="gray",main="",
    xlab=c("","null Wald p-values")[xlab],ylab="",xaxt=c("n","s")[xlab])

  hist(PF[,theta0==0],prob=TRUE,col="gray",main="",
   xlab=c("","null FAB p-values")[xlab],ylab="",xaxt=c("n","s")[xlab])

}

sigNN<-apply(PVALS[,theta0!=0,,]<.05,c(3,4),mean)
sigTN<-apply(PVALS[,theta0==0,,]<.05,c(3,4),mean) 

sigNN

sigTN
```
