---
title: FAB $p$-values for a hidden Markov model
author: Peter Hoff
date: 2019-07-23
---

----

## Summary 
This is a simulation study in which
adaptive FAB $p$-values are constructed using a Gaussian hidden Markov model 
as a linking model. A binary hidden Markov model is the "true" linking model. 
This document serves as the replication code for the example in Section 3.4 of the article
["Smaller $p$-values via indirect information"](https://arxiv.org/abs/XXXX) (Hoff 2019).


Here are some $p$-value functions:
```{r}
#### ---- p-value functions

## -- FAB p-values
pFAB<-function(y,sigma,etheta=0,vtheta=1){
  c(1-abs( pnorm( (y+2*etheta*sigma^2/vtheta)/sigma ) - pnorm(-y/sigma) ))
}

## -- UMPU p-values
pUMPU<-function(y,sigma){
  c(2*pnorm(-abs(y)/sigma ))
}

## -- BH FDR control procedure
rejectBH<-function(pvals,fdr=.1){
  pc<-cbind(sort(pvals),fdr*(1:length(pvals))/length(pvals))
  nd<-min(which( pc[,1]>pc[,2] ) )-1
  order(pvals)[seq(1,nd,length=nd)]
}
```

Here is a function that estimates the parameters in the linking model: 
```{r}
## -- mle from Gaussian hidden Markov model 
mleGHMM<-function(y){
  suppressWarnings( fit<-arima(y,c(1,0,1)) )

  ## TSA4 page 98
  phi<-fit$coef[1] ; theta<-fit$coef[2] ; mu<-fit$coef[3]
  sw2<-fit$sigma2

  v<-sw2*(1+2*theta*phi+theta^2)/(1-phi^2)
  w<-(1+theta*phi)*(theta+phi)/( phi*(1+2*theta*phi+theta^2)  )
  rho<-phi

  sigma2<-v*(1-w)
  psi2<-v*w

  x<-c(sigma2,mu,psi2,rho)
  names(x)<-c("sigma2","mu","psi2","rho")

  gam<-c(x[2]*(1-x[4]),x[4],sqrt(x[3]*(1-x[4]^2)),sqrt(x[1]))
  names(gam)<-c("beta0","beta1","tau","sigma")
  gam
}
```


Here is a function that computes an estimated conditional expectation and variance for the mean at location $j$ given the data at the other locations: 
```{r} 
## -- function to find conditional mean under linking spatial regression model 
cdistGHMM<-function(beta0,beta1,tau,sigma,r){
  w<-2*r+1
  d<-abs(outer(1:w,1:w,"-"))
  C<-(beta1^d)*tau^2/(1-beta1^2)
  iC<-solve(C)
  G<-diag(w)[-(r+1),]
  V<-solve( iC+t(G)%*%G/sigma^2 )
  b0<-c(V%*%( iC%*%(rep(beta0/(1-beta1),w)   ) ))[r+1]
  b1<-(V%*%t(G)%*%G)[r+1,]/sigma^2

  list(b0=b0,b1=b1,vtheta=V[r+1,r+1])
}
```



Here is the simulation study:
```{r,cache=TRUE}
#### ---- HMM parameters 
TP<-rbind( c(.975,.025,.0), c(.01,.99,.01), c(0,.025,.975))
sigma<-1


#### ---- other parameters 
p<-1000
fdr<-.2
r<-50


#### ---- simulation study
EFAB<-XFAB<-UMPU<-pXNULL<-NULL

for(s in 1:100){

  #### ---- generate data
  set.seed(s) 
  theta<-0
  for(j in 2:p){theta<-c(theta,sample(c(-1,0,1),1,prob=TP[theta[j-1]+2,])) } 
  y<-rnorm(p,theta,sigma)

  #### ---- UMPU p-values 
  pU<-pUMPU(y,sigma)
  dU<-rejectBH(pU,fdr) 
  UMPU<-rbind(UMPU,c( sum(theta[setdiff(1:p,dU)]==0), sum(theta[dU]==0),
                      sum(theta[setdiff(1:p,dU)]!=0), sum(theta[dU]!=0) ) )

  #### ---- obtain eBayes means using single estimate of HMM params
  hmm<-mleGHMM(y)
  pripar<-cdistGHMM(hmm[1],hmm[2],hmm[3],hmm[4],r) 
  etheta<-NULL
  for(i in 1:p){
    ii<-seq(i-r,i+r) ; ii[ii<1]<-1 ; ii[ii>p]<-p
    etheta<-c(etheta, pripar$b0 + sum(pripar$b1*y[ii])  )
  }
  pX<-pFAB(y,sigma,etheta,pripar$vtheta)
  dX<-rejectBH(pX,fdr)
  XFAB<-rbind(XFAB,c( sum(theta[setdiff(1:p,dX)]==0), sum(theta[dX]==0),
                      sum(theta[setdiff(1:p,dX)]!=0), sum(theta[dX]!=0) ) ) 
  pXNULL<-c(pXNULL,pX[theta==0] )


  #### ----- exact FAB p-values
  mstheta<-NULL
  for(i in 1:p){ 
    ymi<-replace(y,i,NA)  
    if(i==1){ ymi<-ymi[-1] } 
    if(i==p){ ymi<-ymi[-p] } 
    hmm<-mleGHMM(ymi) 
    pripar<-cdistGHMM(hmm[1],hmm[2],hmm[3],hmm[4],r)
    ii<-seq(i-r,i+r) ; ii[ii<1]<-1 ; ii[ii>p]<-p 
    mstheta<-rbind(mstheta,c(pripar$b0 + sum(pripar$b1*y[ii]),pripar$vtheta ))
  }
  pF<-pFAB(y,sigma,mstheta[,1],mstheta[,2])
  dF<-rejectBH(pF,fdr)
  EFAB<-rbind(EFAB,c( sum(theta[setdiff(1:p,dF)]==0), sum(theta[dF]==0), 
                      sum(theta[setdiff(1:p,dF)]!=0), sum(theta[dF]!=0) )  )
  cat(s,"\n")  
}
```


Save results:
```{r}
colnames(EFAB)<-colnames(XFAB)<-colnames(UMPU)<-
  c("D0E0","D1E0","D0E1","D1E1") 

save(EFAB,XFAB,UMPU,pXNULL,file="resultsHMM.RData")
```


Plot results:
```{r}
## Proportion of discoveries made - power - prob disc given true disc
EPOW<-EFAB[,4]/pmax(1,EFAB[,3]+EFAB[,4])
XPOW<-XFAB[,4]/pmax(1,XFAB[,3]+XFAB[,4])
UPOW<-UMPU[,4]/pmax(1,UMPU[,3]+UMPU[,4])

par(mfrow=c(1,2),mar=c(3,3,1,1),mgp=c(1.75,.75,0))
plot(UPOW,EPOW,xlab="UMPU proportion true effects discovered",
               ylab="FAB proportion true effects discovered")
abline(0,1)

hist(pXNULL,main="",xlab="approximate null p-value",ylab="",prob=TRUE,
     col="gray")
```



